Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-03 appendix.tex
further refinements to appendix D2, D3. plus minor notation update

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the fully-radially-local
approximation, where $\zeta=0$, gives a
(spurious) growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
expected behavior  ($\sigma\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider 
$\beta\ll1$, then setting $\zeta=0$ does not affect the VSI
significantly.

@@

\subsection{Finding marginal stability}\label{disp_neut_limit}
To investigate marginal stability we set $\hat{\sigma} = 0$ so the frequency, $\hat{\upsilon}=\hat{\omega}_c$, is real.  
We take the short radial wavelength limit, $\khat^2\gg 1$, of the coefficients in Eq.\ \ref{coeffs}.  The real and 
imaginary parts of Eq. \ref{relax_disp} then give relations for $\hat{\omega}_c$ and $\beta_c$, the cooling time for 
marginal stability, as 
\begin{align}\
  0 =&M(M+1)(1-h^2 q^2 \khat^2) + 4\beta_c h q M(M+1)\left(\hat{\omega}_c\khat\right) 
 +\left[1 +
    \gamma\beta_c^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}_c\khat\right)^2 \label{relax_cond1} \\
 &+ 2h q \gamma\beta_c \left(\hat{\omega}_c\khat\right)^3 -  \left(\hat{\omega}_c\khat\right)^4 
  + \beta_c^2\gamma^2\hat{\omega}_c^2
  \left(\hat{\omega}_c\khat\right)^4, \notag \\
   0=& 2h q M (M+1)\khat +
   \beta_c(1-\gamma)(1+2M)^2\khat\left(\hat{\omega}_c\khat\right) 
   + h q \khat \left(\hat{\omega}_c\khat\right)^2 -
   2\beta_c\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^2
   - h q
   \gamma^2\beta_c^2\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^3 \label{relax_cond2} \\
   &+
   2\beta_c\gamma\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^4 \notag. 
\end{align}
We recall that in the low frequency and thin-disk approximations, $h,
|\hat{\omega}_c| \ll 1$. We note that $|q|=O(1)$ and 
$\gamma>1$ but is $O(1)$. We consider $M$ to be $O(1)$ or smaller, so 
that $|\hat{\omega_c}\khat|=O(1)$, since for inertial waves
$|\hat{\omega}\khat|^2\simeq 1 + M $. Finally, we assume 
$\beta_c \ll 1$, to be justified \emph{a posteriori}. Applying these
approximations give
Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}.

\subsection{Marginal stability for the fundamental mode}\label{bcrit_alt}
If we apply the above approximations for the fundamental mode $M=0$ to
Eq. \ref{relax_cond1}---\ref{relax_cond2}, but with the less
restrictive assumption $\beta_c \lesssim O(1)$, we expect the balance  
\begin{subequations}\begin{align}
    &0  = \left[1 + \gamma\beta_c^2\left(1-\gamma\right)\right] + 2h q
    \gamma \beta_c \left(\hat{\omega}_c\khat\right) - \left(\hat{\omega}_c\khat\right)^2,\\
    &0 =  \beta_c(1-\gamma) + hq \left(\hat{\omega}_c\khat\right). 
  \end{align}\end{subequations}
Then
\begin{align}
  \beta_c = \frac{h|q|}{\gamma-1}\left[1 - \frac{\gamma h^2 
      q^2}{\gamma-1}\right]^{-1/2},
\end{align}
which reduces to Eq. \ref{iso_vsi_cond} for thin-disks ($h\ll1$). We
thus deduce that in fact $\beta_c\ll 1$, which motivates us to make this
strong assumption from the onset.

>>>>

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the fully-radially-local
approximation, where $\zeta=0$, gives a
(spurious) growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
expected behavior  ($\sigma\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider 
$\beta\lesssim O(1)$, then setting $\zeta=0$ does not affect the VSI
significantly.

@@

\subsection{Finding marginal stability}\label{disp_neut_limit}
To investigate marginal stability we set $\hat{\sigma} = 0$ so the
frequency, $\hat{\upsilon}=\hat{\omega}_c$, is real; and $\beta =
\beta_c$,  the cooling time for marginal stability. 
We take the short radial wavelength limit, $\khat^2\gg 1$, of the
coefficients in Eq.\ \ref{coeffs}. We consider $M \lesssim O(1)$ and
assume $\beta_c \ll \khat$.  The real and 
imaginary parts of Eq. \ref{relax_disp} then give relations for $\hat{\omega}_c$ and $\beta_c$ as 
\begin{align}
  0 =&M(M+1)(1-h^2 q^2 \khat^2) + 4\beta_c h q M(M+1)\left(\hat{\omega}_c\khat\right) 
 +\left[1 +
    \gamma\beta_c^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}_c\khat\right)^2 \label{relax_cond1} \\
 &+ 2h q \gamma\beta_c \left(\hat{\omega}_c\khat\right)^3 -  \left(\hat{\omega}_c\khat\right)^4 
  + \beta_c^2\gamma^2\hat{\omega}_c^2
  \left(\hat{\omega}_c\khat\right)^4, \notag \\
   0=& 2h q M (M+1)\khat +
   \beta_c(1-\gamma)(1+2M)^2\khat\left(\hat{\omega}_c\khat\right) 
   + h q \khat \left(\hat{\omega}_c\khat\right)^2 -
   2\beta_c\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^2
   - h q
   \gamma^2\beta_c^2\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^3 \label{relax_cond2} \\
   &+
   2\beta_c\gamma\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^4 \notag. 
\end{align}
We recall that in the low frequency and thin-disk approximations, $h,
|\hat{\omega}_c| \ll 1$. We note that $|q|=O(1)$ and 
$\gamma>1$ but is $O(1)$. We assume $|\hat{\omega_c}\khat|=O(1)$,
since for inertial waves $|\hat{\omega}\khat|\simeq \sqrt{1 + M}$. Finally, we further assume 
$\beta_c \ll 1$, to be justified \emph{a posteriori}, to give 
Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}.

\subsection{Marginal stability for the fundamental
  mode}\label{bcrit_alt}
For the fundamental mode ($M=0$) we can deduce $\beta_c\ll 1$, which
motivates us to make this strong assumption from the onset. In this
case, Eq. \ref{relax_cond1} --- \ref{relax_cond2} 
imply the following balance for large $\khat$ and small
$\hat{\omega_c}$ while $\hat{\omega}_c\khat$ remain fixed:  
\begin{subequations}\begin{align}
    &0  = \left[1 + \gamma\beta_c^2\left(1-\gamma\right)\right] + 2h q
    \gamma \beta_c \left(\hat{\omega}_c\khat\right) - \left(\hat{\omega}_c\khat\right)^2,\\
    &0 =  \beta_c(1-\gamma) + hq \left(\hat{\omega}_c\khat\right). 
  \end{align}\end{subequations}
Then
\begin{align}
  \beta_c = \frac{h|q|}{\gamma-1}\left(1 - \frac{\gamma h^2 
      q^2}{\gamma-1}\right)^{-1/2},
\end{align}
which reduces to Eq. \ref{iso_vsi_cond} at $O(h)$.
